### 2017-10-09 response/response-jde-1.tex

[also speaks to whether this is a conscious decision or not] 
The table also supports our conjecture that the estimated effects 
[might need a less causal word] 
of crop loss capture a conscious decision. 
If shocks lead to lower probability of pregnancy and birth because
of other factors than the decision to use contraceptive we should see
a negative and statistically significant coefficient for the crop loss
dummy by itself.
We do not.
Instead, what we get is a substantial negative coefficient for the interaction
between contraceptive use and crop loss and a coefficient for crop loss 
on its own that is positive and far from statistically significant.

\item Several appendix tables interact crop loss with initial assets in
Round 1 of the survey. This exercise is interesting and worthy of doing,
given that one might expect asset-poor households to have particular
difficulty in smoothing consumption. But the authors should leave Round
1 outcomes out of this regression.

[RESPONSE:] We have added the results where round 1 shocks has been 
excluded to the relevant table in the Appendix section on age
groups, wealth, and education.
The results are broadly consistent with what we originally found
for pregnancy and birth, 
The contraceptive use results have the correct sign but the direct
effect is not statistically significant, presumably because of the 
lower variation in crop loss and the smaller sample size.

### 2017-10-09 response/response-jde-1.tex

\begin{center} Response to Referee Comments \end{center}

\noindent Please find attached the revised version of our paper,
``Income Shocks, Contraceptive Use, and Timing of Fertility.''
We would like to thank the two referees for the very useful comments and
suggestions,
which we believe has improved the paper.

We rewrote the code for cleaning and analyzing the data from the ground up 
to make it easier to follow and eventually publish on GitHub.
This rewrite lowered the sample by three observations.
Including all married/partnered women up to age 45, rather than exclude
women younger than 18, the final sample size is, however, the same as 
before and the results are essentially unchanged.

To remain within the page limit we had to substantially reorganize
part of the paper.
The following is a list of the more substantial changes made:
\begin{enumerate}
\item We removed the map of Kagera.
\item In response to the reviewers' comments we have almost completely
rewritten the Robustness section as detailed below.
\item As part of the rewrite of the Robustness section we moved 
the discussion of age, wealth, and education results to a new Appendix
section since these results do little to clarify the mechanisms 
behind the main results.
We still address and incorporate the comments from reviewers that
relate to these outcomes.
\item We moved the discussion of estimation strategy for age,
wealth, and education estimation to the new Appendix section.
\item We have added another new Appendix section that examine results 
with including an additional lag for crop loss (see response to
Reviewer 1 below).
\item A number of new Appendix tables have been added in response
to various comments and other edited.
\end{enumerate}

We have tried to highlight the factors that can affect our identification 
strategy, but if we need to tone down the causality claims further
we are happy to do so.
We list below our responses to the individual comments and suggestions.

### 2017-10-09 response/response-jde-1.tex

% This interesting paper analyzes the relationship between income shocks
% (as proxied by self-reported crop loss) and marital fertility, using a
% small Tanzanian panel survey from the early 1990s. A key feature of the
% paper is that the authors are able to observe use of both traditional
% and modern birth control techniques. In regression specifications with
% and without community or woman fixed effects, the authors find that crop
% loss is associated with decreases in the probability of conceiving, with
% a corresponding increase in use of traditional birth control techniques
% but not modern contraceptives.

The paper could benefit from focusing its motivation somewhat more. An
important point (that the authors do not emphasize enough) is that use
of modern contraception in the sample is extremely low, so the lack of a
relationship between shocks and use of modern methods is not surprising.
The more noteworthy result is that use of traditional methods responds
to shocks. In this sense, the paper makes an important contribution to
understanding how populations control fertility before widespread
adoption of modern contraception. For instance, it is well known that
marital fertility in Europe declined largely before modern contraception
became available, but the methods couples used to achieve this reduction
are not well documented (see, for example, Guinnane JEL 2011). The paper
would be even more compelling if the authors focused the paper's
motivation on the questions of whether and how couples modify fertility
to respond to changing economic conditions before the advent of
modern contraception. That motivation would also help justify their
choice of dataset, which otherwise seems somewhat small and out of date
relative to other available panel datasets from developing countries.

[RESPONSE:]

We have edited the Introduction to highlight the role of traditional 
contraception and that most of the increase in contraception after a 
shock comes from traditional methods.
We have also added an additional point to our list of the main 
contributions of the paper reflecting your suggestion.
In addition to the \citet{Guinnane2011} paper, we also cite a just
published paper on spacing in pre-fertility transition England
\citep{Cinnirella2017} and \citet{Michael1976}, \citet{David1986},
and \citet{Santow1995}
on the role of different types of traditional contraception in
controlling fertility before the emergence of modern contraception.

### 2017-10-09 paper/incomeShocks-jde-r1.tex
Respond to R1 on the role of traditional contraceptives

% [Lead with this or the next sentence?]
% Our result that people are able to postpone fertility using mainly 
% traditional contraceptive do, however, have implications for our 
% understanding of long-term fertility.
% [Alternative lead]
What implications do our results have for our understanding of 
long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.%
\footnote{
See, for example, \cite{Portner2014a} and the literature surveyed therein.
}
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.

>>>>

% [Lead with this or the next sentence?]
% Our result that people are able to postpone fertility using mainly 
% traditional contraceptive do, however, have implications for our 
% understanding of long-term fertility.
% [Alternative lead]
What implications do our results have for our understanding of 
long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.%
\footnote{
See, for example, \citet{Portner2014a} 
and \citet{Portner2018} and the literature surveyed therein.
}
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.
